---
title: "CA1"
author: "Florence Lourdes"
date: "2023-09-27"
output: html_document
statement of purpose: "I am training a model to predict the price range of houses depending on some variables. 
I am doing it for people who are looking to buy a house or sell their house. They can benefit by getting information about what they should be expecting when it comes to certain price range houses."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#install.packages("tidyverse")
#install.packages("corrr")
library(corrr)
library(tidyverse)
library(tidymodels)
library(kableExtra)
library(patchwork)
library(ggplot2)
#install.packages('xgboost')
require(xgboost)
data(agaricus.train, package='xgboost')
data(agaricus.test, package='xgboost')
train <- agaricus.train
test <- agaricus.test
data <- read_csv("C:/MAT434/data.csv")
```
# Head of Data
The following table shows the top 6 rows of the data
```{r}
data %>%
  head()
```

# Set a seed for reproducibility
Setting a seed make sure that we get the same results for randomization
```{r}
set.seed(123)
```

# Split data into training and test sets
Splitting data into a training and test sets with a 8 to 2 ratio and marking price range as prediction target 
```{r}
split_data <- initial_split(data, prop = 0.8, strata = priceRange)  # Adjust ratio and strata as needed
```

# Extract the training and test sets
Extracting the data from each training and test sets. These patterns include outliers and features of the data that might be unexpected. EDA is an important first step in any data analysis
```{r}
train_data <- training(split_data)
test_data <- testing(split_data)
```

# Build cross-validation folds
K-fold cross-validation is a technique for evaluating predictive models. The dataset is divided into k subsets or folds. The model is trained and evaluated k times, using a different fold as the validation set each time. Performance metrics from each fold are averaged to estimate the model's generalization performance.
```{r}
set.seed(456)
train_folds <- vfold_cv(train_data, v = 10)
```

# Exploratory Data Analysis (EDA)
EDA is an analysis approach that identifies general patterns in the data. These patterns include outliers and features of the data that might be unexpected. EDA is an important first step in any data analysis.
```{r}
# Summary statistics
summary(train_data)

# Bar plots for categorical variables
ggplot(train_data, aes(x = priceRange, y=numOfBedrooms)) + 
  geom_bar(stat = "identity")

# Scatterplots for relationships
ggplot(train_data, aes(x=priceRange, y=numOfBathrooms)) + 
    geom_point()

# Correlation matrix
data_cor <- train_data %>% 
  select(numOfBedrooms, lotSizeSqFt, priceRange) %>% 
  correlate()
```

# Training a Model

```{r}
#Create a model specification
log_reg_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

#Create a recipe
log_reg_rec <- recipe(priceRange ~ ., data = train_data) %>%
  step_rm(id, description, latitude, longitude, avgSchoolRating, MedianStudentsPerTeacher) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) #add recipe steps as needed

#Create a workflow
log_reg_wf <- workflow() %>%
  add_model(log_reg_spec) %>%
  add_recipe(log_reg_rec)

#Run cross-validation to obtain cross-validation performance estimate
log_reg_cv_results <- log_reg_wf %>%
  fit_resamples(train_folds)

#Collect cross-validation results
#log_reg_cv_results %>%
  #collect_metrics()

#Fit model to training data
#log_reg_fit <- log_reg_wf %>%
  #fit(train_data)

#Assess model on test data
#log_reg_wf %>%
  #augment(test) %>%
  #my_metrics(response, .pred)

#Use model to predict for new data
#log_reg_fit %>%
  #augment(new_data)
```