---
title: "Halloween Comptetition"
author: "Florence Lourdes"
date: "2023-10-31"
output: html_document
---

```{r}
##install.packages("caret")
library(rpart.plot)
library(rpart)
library(tidyverse)
library(tidymodels)
library(caret)
library(kknn) # For K-nearest neighbors

monster_train <- read_csv("C:/MAT434/MAT434 R Studio/monster_train.csv")
monster_test <- read_csv("C:/MAT434/MAT434 R Studio/monster_comp.csv")
```
# Build cross-validation folds
K-fold cross-validation is a technique for evaluating predictive models. The dataset is divided into k subsets or folds. The model is trained and evaluated k times, using a different fold as the validation set each time. Performance metrics from each fold are averaged to estimate the model's generalization performance.
```{r}
set.seed(456)
train_folds <- vfold_cv(monster_train, v = 10)
```

# Model Construction and Interpretation
In the following code, we will be using two training models: nearest neighbor and decision tree.The steps are to make a workflow, evaluate, and print the result.

```{r}
#knn workflow
knn_spec <- nearest_neighbor() %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_rec <- recipe(class ~ height + color + rottingFleshPct,bloodCoverage, aura, hairLength, data = monster_train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_wf <- workflow() %>%
  add_model(knn_spec) %>%
  add_recipe(knn_rec)

#decision tree workflow
dt_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

dt_rec <- recipe(class ~ height + color + rottingFleshPct,bloodCoverage, aura, hairLength, data = monster_train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

dt_wf <- workflow() %>%
  add_model(dt_spec) %>%
  add_recipe(dt_rec)

#evaluate workflows
knn_cv_results <- knn_wf %>%
  fit_resamples(train_folds)

dt_cv_results <- dt_wf %>%
  fit_resamples(train_folds)
```

```{r}
#knn result
knn_cv_results %>%
  collect_metrics() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```


```{r}
#dt result
dt_cv_results %>%
  collect_metrics() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

```{r}

##grid
grid_neighbors <- tibble(
  "neighbors" = c(1, 3, 5, 7, 9, 11, 15, 21, 27, 30, 36, 41, 50)
  )

grid_depth <- tibble(
  "tree_depth" = c(2, 3, 4, 5, 8, 10, 12, 15, 20)
  )
```

```{r}
##knn workflow
knn_spec <- nearest_neighbor(neighbors = tune()) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_rec <- recipe(class ~ height + color + rottingFleshPct,bloodCoverage, aura, hairLength, data = monster_train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_wf <- workflow() %>%
  add_model(knn_spec) %>%
  add_recipe(knn_rec)

##dt workflow
dt_spec <- decision_tree(tree_depth = tune()) %>%
  set_engine("rpart") %>%
  set_mode("classification")

dt_rec <- recipe(class ~ height + color + rottingFleshPct,bloodCoverage, aura, hairLength, data = monster_train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

dt_wf <- workflow() %>%
  add_model(dt_spec) %>%
  add_recipe(dt_rec)

##hyperparameter tune
knn_tune_results <- knn_wf %>%
  tune_grid(
    grid = grid_neighbors,
    resamples = train_folds
  )

dt_tune_results <- dt_wf %>%
  tune_grid(
    grid = grid_depth,
    resamples = train_folds
  )

```

Here is the nearest neighbor result:
```{r}
##knn result 
knn_tune_results %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(neighbors, mean, std_err) %>%
  arrange(-mean) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

Here is the decission tree result:
```{r}
##dt results
dt_tune_results %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  select(tree_depth, mean, std_err) %>%
  arrange(-mean) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

```{r}
knn_spec <- nearest_neighbor(neighbors = 50) %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_rec <- recipe(class ~ height + color + rottingFleshPct,bloodCoverage, aura, hairLength, data = monster_train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_wf <- workflow() %>%
  add_model(knn_spec) %>%
  add_recipe(knn_rec)

knn_fit <- knn_wf %>%
  fit(monster_train)
```

```{r}
my_submissions <- knn_fit %>%
  augment(monster_test) %>%
  select(id, .pred_class) %>%
  rename(class = .pred_class)
write.csv(my_submissions, "submission.csv", row.names = FALSE)
```