---
title: "The reduction in weekly working hours in Europe" 
subtitle: "Looking at the development between 1996 and 2006"
author: "Florence Lourdes"
output: html_document
---

## Summary 

The **International Labour Organization (ILO)** has many [data sets](http://www.ilo.org/global/statistics-and-databases/lang--en/index.htm) on working conditions. For example, one can look at how weekly working hours have been decreasing in many countries of the world, while monetary compensation has risen. In this report, *the reduction in weekly working hours* in European countries is analysed, and a comparison between 1996 and 2006 is made. All analysed countries have seen a decrease in weekly working hours since 1996 – some more than others.

## Preparations 

```{r loading_packages, message = FALSE}
library(dplyr)
library(ggplot2)
library(forcats)
```

## Analysis

### Data

The herein used data can be found in the [statistics database of the ILO](http://www.ilo.org/ilostat/faces/wcnav_defaultSelection;ILOSTATCOOKIE=ZOm2Lqrr-OIuzxNGn2_08bNe9AmHQ1kUA6FydqyZJeIudFLb2Yz5!1845546174?_afrLoop=32158017365146&_afrWindowMode=0&_afrWindowId=null#!%40%40%3F_afrWindowId%3Dnull%26_afrLoop%3D32158017365146%26_afrWindowMode%3D0%26_adf.ctrl-state%3D4cwaylvi8_4). For the purpose of this course, it has been slightly preprocessed.

```{r loading_data}
load(url("http://s3.amazonaws.com/assets.datacamp.com/production/course_5807/datasets/ilo_data.RData"))
```

The loaded data contains `r ilo_data %>% nrow()` rows and `r ilo_data %>% ncol` columns. 

```{r generating_summary_statistics}
# Some summary statistics
ilo_data %>%
 group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours))
```

```{r}
summary <- ilo_data %>%
 group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours))
```

```{r}
hours <- summary %>%
  select(mean_working_hours)
compensation <- summary %>% 
  select (mean_hourly_compensation)
```

As can be seen from the above table, the average weekly working hours of European countries have been descreasing to `r hours summary %>% tail()` since 1980.

While the mean hourly compensation has increased to `r compensation %>% tail()` since 1980.

### Preprocessing

The data is now filtered so it only contains the years 1996 and 2006 – a good time range for comparison. 

```{r}
ilo_data <- ilo_data %>%
  filter(year == "1996" | year == "2006")
  
# Reorder country factor levels
ilo_data <- ilo_data %>%
  # Arrange data frame first, so last is always 2006
  arrange(year) %>%
  # Use the fct_reorder function inside mutate to reorder countries by working hours in 2006
  mutate(country = fct_reorder(country,
                               working_hours,
                               last))
```  

### Results

In the following, a plot that shows the reduction of weekly working hours from 1996 to 2006 is produced.

```{r}
ilo_data %>%
  group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours))
x <- c(33.15975, 31.78460)
```

The result is that there is a `r diff(x)` difference of working hour between 1996 and 2006 